---
title: "MSDS 6303 - Case Study 2"
author: "Chayson Comfort, Phillip Edwards, George Sturrock"
date: "August 10, 2017"
output: 
  md_document:  
    variant:  markdown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Introduction

This presentation is composed of two main sections.  The section titled "Question 2" analyzes orange tree growth.  The "Question 3" section analyzes temperature trends on a country and city bases.  The procedure to load, cleanse and analyze data will be described in each section.

## Libraries

Three R libraries were used to address the requirements for both sections.  "ggplot2" is used for all visualizations.  "dplyr" is used to cleanse and manipulate data. "gridExtra" is used specifically to produce a combined plot in the final section of this document.

```{r libraries}
source("CS2Libraries.R")
print("Libraries Loaded")
```

## Question 2

The data set named "Orange" is delivered with the "datasets" R package.  The "Orange" data set includes three columns:  Tree, Age and Circumference.  This data is used to calculate the mean and median of the trunk circumferences for the different trees, plot the tree circumference agains the age and to produce a box plot to show circumference data by tree.  This data set required no cleansing prior to analysis.

The code for cleansing and analyzing the "Orange" data set is sourced in the code chunk below:
```{r q2source}
source("CS2Q2Analysis.R", print.eval = FALSE)
print("Question 2 Analysis Complete")
```

The "Orange" data set is loaded into R for analysis.
```{r q2data, eval=FALSE}
data("Orange")
```

### Question 2.a
The mean and median for the entire Orange tree population is calculated.
```{r q2a1}
cat("The median circumference for all tree measurements is", median(Orange$circumference),".", "The mean circumference for all tree measurements is", mean(Orange$circumference),".")
```

The mean and median circumference of each tree is calculated.
```{r q2a2}
#TreeCirc <- Orange %>% group_by(Tree) %>% summarise(circmedian=median(circumference), circmean=mean(circumference))
print(TreeCirc)
```

### Question 2.b
Two plots are produced to show the relationship between tree circumference and age.  The first plot is a scatter plot mapping the relationship between tree age and circumference.  The data shows tree circumference increases as each tree ages.  The second plot is the contains lines connecting the data points.  Adding lines makes it easier to see the positive relationship between tree age and circumference.
```{r q2b}
ggplot(Orange, aes(x=circumference, y=age, shape=Tree, color=Tree)) +geom_point() +ggtitle("Orange Tree Circumference vs Age")

ggplot(Orange, aes(x=circumference, y=age, shape=Tree, color=Tree)) +geom_point() +geom_line() +ggtitle("Orange Tree Circumference vs Age with Lines")
```

### Question 2.c
A box plot is created to show circumference data by tree.
```{r q2c}
ggplot(Orange, aes(x=Tree, y=circumference, fill=Tree)) +geom_boxplot() +ggtitle("Tree Circumference")
```

## Question 3
This section utilizes two data sets to analyze historical temperature data for countries and cities.  The two data sets are in CSV format, stored locally and loaded into R using read.csv for cleansing and analysis.
```{r q3loadcode, eval=FALSE}
dataloc <- paste(getwd(),"/data/TEMP.csv", sep = "")
print(dataloc)
TempData <- read.csv(dataloc, stringsAsFactors = FALSE, header = TRUE)
dataloc <- paste(getwd(),"/data/CityTemp.csv", sep = "")
print(dataloc)
CityTemp <- read.csv(dataloc, stringsAsFactors = FALSE, header = TRUE)
```

The following chunk executes the data load procecures shown above.
```{r q3load}
source("CS2Q3Load.R", print.eval = FALSE)
```

### 3.i
This section utilizes the "TempData" data frame created above to find the top 20 countries with the maximum difference between the monthly maximum average and minimum average temperatures since 1900.  

A subset of the "TempData" data frame is created to store temperature data since 1900, and the date column is coerced into date format.  
```{r q3i1, eval=FALSE}
TempData$First2ofDate <- substr(TempData$Date, 1, 2)
TempData$First2ofDate <- as.integer(TempData$First2ofDate)
GT1899TempData <- subset(TempData, (First2ofDate < 17 | is.na(First2ofDate) == TRUE) & complete.cases(Monthly.AverageTemp)== TRUE)
GT1899TempData$Date <- as.Date(GT1899TempData$Date, "%m/%d/%Y")
```

Next, the maximum and minimum monthly averages by country are calculated and stored in a new data frame.  The difference between the maximum and minimum is then calculated and stored in a new colum.  That data is sorted and subset to arrive at the top 20 countries by difference between maximum and minimum monthly average since 1900.
```{r q3i2, eval=FALSE}
#Find max and min monthly avg temp by country
MMCTempData <- GT1899TempData %>% group_by(Country) %>% summarise(MaxMonthlyAvg=max(Monthly.AverageTemp), MinMonthlyAvg=min(Monthly.AverageTemp))

#Calculate Difference in Max and Min Average Temp by Country
MMCTempData$Diff <- (MMCTempData$MaxMonthlyAvg - MMCTempData$MinMonthlyAvg)

#Sort by Diff Descending
MMCTempData <- MMCTempData[order(MMCTempData$Diff, decreasing = TRUE),]

#Subset the top 20
MMCTop20 <- MMCTempData[1:20,]
```

The code is executed below.
```{r q3isource}
source("CS2Q3i.R", print.eval = FALSE)
```

The following scatter plot shows the top 20 countries by difference between monthly maximum and minimum since 1900.
```{r q3iplot}
ggplot(MMCTop20, aes(x=Diff, y=reorder(Country, Diff))) +geom_point() +labs(x="Temperature Difference", y="Country") +ggtitle("Top 20 Countries", subtitle="by difference between Max and Min Monthly Average")
```

### 3.ii
The "TempData" data frame is next utilized to analyze United States temperature data.  The "TempData" data frame is subset to include only United States temperature data since 1990, a new column for Year is created and the monthly average temperature in degrees Celcius is used to create a new column store monthly average temperature in degrees Fahrenheit.

```{r q3iisource}
source("CS2Q3ii.R", print.eval = FALSE)
```

```{r q3ii1, eval=FALSE}
USTemp <- subset(GT1899TempData, Country == 'United States')

#Create average land temp by year, subset for for dates greater than 1989
USTemp$Year <- as.numeric(format(USTemp$Date, "%Y"))
USTemp <- subset(USTemp, Year > 1989)

#Create (°F) column
USTemp$Monthly.AverageTemp.F <- (USTemp$Monthly.AverageTemp * (9/5) + 32)
```

The second requirement of this question is to produce the mean of the United States monthly average temperatures by year.     
```{r q3iiyoy, eval=FALSE}
#Calculate average yearly temperature and create new dataframe
USTempYear <- USTemp %>% group_by(Year) %>% summarise(MeanTempF=mean(Monthly.AverageTemp.F))
```

The result is plotted in a scatter plot.  A linear regression line is added to the scatter plot to clearly show yearly temperature trends over time.
```{r q3iiplot}
#Plot yearly temperature averages and Calculate linear regression for input into ggplot abline function
reg1 <- lm(USTempYear$MeanTempF ~ USTempYear$Year)
ggplot(USTempYear, aes(x=Year, y=MeanTempF, group=1)) +geom_point() +labs(y="Average Temperature (°F)") +geom_abline(intercept = reg1$coefficients[1], slope = reg1$coefficients[2], color = "red", linetype = "dashed", size = 1) +ggtitle("Average US Land Temperature by Year")
```

Finally, year over year temperature differences are calculated and the greatest year over year difference is identified.
```{r q3iic, eval=FALSE}
#Calculate Year Over Year average temparature difference
USTempYearYOY <- USTempYear %>% mutate(YoY = MeanTempF - lag(MeanTempF), PriorYear = lag(Year))
USTempYearYOY <- USTempYearYOY[order(USTempYearYOY$YoY, decreasing = TRUE),]
```
```{r q3iiic1}
#Print maximum year over year change with the corresponding years
cat("The maximum Year over Year Temperature Difference is:",USTempYearYOY$YoY[1], "This difference occured in", USTempYearYOY$Year[1], "The prior year was", USTempYearYOY$PriorYear[1])
```

### 3.iii
The objective of section 3.iii is similar to 3.i.  Instead of analyzing temperature data at the country level.  This section uses the "CityTemp" data frame to find the difference between the maximum and minimum monthly average temperatures by city.  This data is then sorted, the top 20 cities by greatest temperature difference are identified and plotted on a scatter plot.
```{r cs2q3iii}
#Standardize dates prior to subsetting data
CityTemp$First2ofDate <- substr(CityTemp$Date, 1, 2)
CityTemp$First2ofDate <- as.integer(CityTemp$First2ofDate)
CityTempGT1899 <- subset(CityTemp, (First2ofDate < 17 | is.na(First2ofDate) == TRUE) & complete.cases(Monthly.AverageTemp)== TRUE)
CityTempGT1899$Date <- as.Date(CityTempGT1899$Date, "%m/%d/%Y")

#Find max and min monthly avg temp by country
CityTempMaxMin <- CityTempGT1899 %>% group_by(Country, City) %>% summarise(MaxMonthlyAvg=max(Monthly.AverageTemp), MinMonthlyAvg=min(Monthly.AverageTemp))

#Calculate Difference in Max and Min Average Temp by Country
CityTempMaxMin$Diff <- (CityTempMaxMin$MaxMonthlyAvg - CityTempMaxMin$MinMonthlyAvg)

#Sort by Diff Descending
CityTempMaxMin <- CityTempMaxMin[order(CityTempMaxMin$Diff, decreasing = TRUE),]

#Subset the top 20 and concatenate City and Country
CityTempTop20 <- CityTempMaxMin[1:20,]
CityTempTop20$CityCountry <- paste(CityTempTop20$City, ", ", CityTempTop20$Country, sep = "")

#Plot the top 20
ggplot(CityTempTop20, aes(x=Diff, y=reorder(CityCountry, Diff))) +geom_point() +labs(x="Temperature Difference", y="City") +ggtitle("Top 20 Cities", subtitle="by difference between Max and Min Monthly Average")
```

### 3.iv
Sections 3.i and 3.iii show similar plot graphs of the Top 20 Countries and Cities by difference between monthly average maximum and minimum temperature differences.  The greatest difference is the absense of China from the Country data set.  Cities in China occupy nine of the twenty top cities on the city plot.  It is possible China would be the top on the country plot if included in the country data set.  

Additionally, these data sets show the most monthly average temperature differences can be found on the Asian continent.  Approximately half of the countries and 75% of the cities in the respective top 20 lists are located in Asia.
```{r csq3iiii}
countryplot <- ggplot(MMCTop20, aes(x=Diff, y=reorder(Country, Diff))) +geom_point() +labs(x="Temperature Difference", y="Country") +ggtitle("Top 20 Countries", subtitle="by difference between Max and Min Monthly Average")

cityplot <- ggplot(CityTempTop20, aes(x=Diff, y=reorder(CityCountry, Diff))) +geom_point() +labs(x="Temperature Difference", y="City") +ggtitle("Top 20 Cities", subtitle="by difference between Max and Min Monthly Average")

grid.arrange(countryplot, cityplot, ncol=2)
```
